Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-06-23 doc/thesis/methods_results.Rnw
Most thesis changes done

As described in \cref{sec:smc}, \software{netabc} keeps track of a population
of particles $x^{(k)}$ indexed by an integere $k$, each of which contains
particular parameter values $\theta^{(k)}$ for the \add{contact network} model
we are trying to fit \add{to the input tree}. A small number \gls{M} of contact
networks $z^{(k,i)}$, $1 \leq i \leq \gls{M}$, are generated under the model
for each particle in accordance with that particle's parameters. An epidemic is
simulated over each of these networks using Gillespie simulation, and by
keeping track of its progress, a transmission tree is obtained. Thus, each
particle becomes associated with several simulated transmission trees. These
trees are compared to the input tree using the tree kernel. Particles are
weighted according to the similarity of their associated simulated trees with
the true tree, with more similar trees receiving higher weights. The particles
are iteratively perturbed to explore the parameter space, and particles with
simulated trees too distant from the true tree are periodically dropped and
resampled. Once a convergence criterion is attained, the final set of particles
is used as a Monte Carlo approximation to the target distribution of \gls{ABC},
which is assumed to resemble the posterior distribution on model parameters
(see \cref{sec:abc}). A graphical schematic of this algorithm is shown in
\cref{fig:abcsmc}.

@@

\add{We present here a non-rigorous justification for the use of
\gls{ABC} for the problem at hand, as opposed to more frequently-used
approaches for fitting mathematical models (see \cref{chp:prelim}). Consider a
contact network model with parameters $\theta$, and an estimated transmission
tree \gls{T}. Taking a Bayesian approach, our aim is to obtain a sample from
the posterior distribution on the model's parameters given our data,}
\[
    \madd{\post(\theta \mid \gls{T}) = \frac{\lik(\gls{T} \mid \theta) \prior(\theta)}
    {\int_{\Theta}\lik(\gls{T} \mid \theta) \prior(\theta) \d\theta}.}
\]
\add{For all but the simplest models, the normalizing constant in the
denominator is an intractable integral. What we shall argue here is that, in
contrast to most commonly studied mathematical models, the likelihood
$\lik(\gls{T} \mid \theta)$ is also likely to be intractible in our case.}

@@

\add{We have shown that both the normalizing constant $\int_\Theta \lik(\gls{T}
\mid \theta) \prior(\theta) \d\theta$ and the likelihood $\lik(T \mid \theta)$
are likely computationally prohibitive to calculate. If this is the case, the
problem of fitting contact network models to phylogenies seems to be of the
\emph{doubly-intractible} type~\autocite{murray2012mcmc}, which would imply that
these models are not amenable to neither \gls{ML} nor Bayesian inference
techniques. Although both methods are able to cope with an intractible
normalizing constant (for example, by local search for \gls{ML} or Bayesian
\gls{MCMC}), neither can avoid the intractable likelihood calculations. This
justifies the use of \gls{ABC}, which is a likelihood-free method.}

@@

\add{From a public health standpoint, all four parameters are of some interest.
The prevalence \gls{I} can be used to estimate the resources required to combat
an ongoing epidemic, while total susceptible population size \gls{N} provides a
similar metric for preventative measures. The average degree of the network, in
this case $2\gls{m}$, is directly related to \gls{R0}, the basic reproductive
number~\autocite{britton2002bayesian}. \gls{R0} quantifies the number of
secondary infections ultimately caused by one infected individual; higher
\gls{R0} generally indicates faster epidemic growth and/or larger eventual
epidemic size~\autocite{anderson1992infectious}. In a homogeneously mixed
population, the proportion of the population which must be vaccinated to
control an epidemic can be expressed in terms of
\gls{R0}~\autocite{anderson1982directly}. Although optimal vaccination
strategies may differ in heterogeneous contact
structures~\autocite{ma2013importance}, it is reasonable to suppose that there
would still be a relationship between \gls{m}, \gls{R0}, and the vaccination
threshold. The preferential attachment power \gls{alpha} quantifies the degree
to which high-degree nodes, also called
superspreaders~\autocite{kemper1980identification}, characterize the network
structure. Superspreaders have been hypothesized to play a greater-than-average
role in the spread of several diseases~\autocite{stadler2013uncovering,
shen2004superspreading}. If so, network-based
interventions~\autocite{little2014using, wang2015targeting} may be worth
considering as part of an epidemic control strategy. \gls{alpha} can also offer
some insight into how the network would react to the removal of nodes.
\textcite{dombrowski2013topological} found evidence of preferential attachment
in \gls{IDU} networks, and suggested that as a consequence of this
characteristic, the removal of random nodes (such as through a police
crackdown) might inadventently make it easier for epidemics to spread. When
individuals with only one or two connections lose them, they might tend to seek
out well-known (that is, high-degree) members of the community, thus increasing
those individuals' connectivity even further.}

\add{All four \gls{BA} parameters can be estimated using more conventional
approaches, but this estimation may be challenging depending on characteristics
of the epidemic under consideration. In theory, any network parameter can be
estimated by explicitly constructing the contact network, although this is
highly resource intensive and is hampered by misreporting and other
challenges~\autocite{eames2015six}. All parameters become more difficult to
estimate when the infected population is ``hidden'' due to illegal or
stigmatized behaviour, as is sometimes the case with \gls{HIV} outbreaks among
\gls{IDU}, \gls{MSM}, or sex workers. Of course, phylodynamic methods are
equally ineffective when the population is completely hidden, since we must at
least be able to sample viral sequences from its members. If a population is
accessible, \gls{N} and \gls{I} may be estimable by surveying and/or testing
each member of the population for the disease. However, this will not tell us
if there are large compartments of the population that we simply have not
sampled. Our hope is that estimating \gls{N} and \gls{I} phylogenetically might
provide this additional information. The average degree of the network,
$2\gls{m}$, is also estimable by a survey, although individuals may be
unwilling or unable to disclose how many contacts they have had. The estimation
of \gls{alpha} is more complex, as this parameter is most strongly reflected in
the connectivity of very high degree nodes, who are rare in the population.
Locating them might require contact tracing, or respondent-driven
sampling~\autocite{heckathorn1997respondent}. Even if the full degree
distribution of a network is available, there are models other than
preferential attachment which can produce scale-free
networks~\autocite[\eg][]{kumar2000stochastic}. \textcite{de2007preferential}
were able to estimate \gls{alpha} by maximum likelihood using partner count
data from several sequential time intervals, but they admit such detailed data
are not usually available. Moreover, their dataset was constructed via a random
survey, which would likely miss the few high-degree nodes characterizing a
power law degree distribution. In summary, each of the \gls{BA} parameters may
be estimated without phylodynamics, but there are sufficient difficulties that
we believe an alternative method using sequence data is warranted.}

@@

\add{A second experiment was designed to test whether we could actually
estimate the parameters numerically, rather than just telling three different
values apart, and also to assess how identifiability varied in different
regions of the parameter space. An individual tree was compared to simulated
trees on a one-dimensional grid of values of one \gls{BA} parameter, to obtain
a ``distribution'' of kernel score values. From this distribution, estimates
and credible intervals of the parameter could be calculated. Repeating this
experiment with trees located throughout the parameter space allowed us to
beter quantify the identifiability. Furthermore, doing marginal estimation
(that is, estimating one parameter with all others fixed) can provide insight
into any biases observed when doing joint estimation with \gls{ABC}. If grid
search is inaccurate, it indicates a lack of parameter identifiability.
However, if the marginal grid search estimates are accurate but the estimates
obtained with \gls{ABC} are biased, this points to confounding between the
parameters which could only be observed when they are all estimated jointly.}

@@

\add{The Monte Carlo approximation error is fairly easily quantified by simply
increasing the computing power used for \gls{SMC}. We ran one simulation using
a larger number of pakticles, more simulated datasets per particle, and a
higher value for \gls{alphaess}. A substantial improvement in accuracy
resulting from these changes would likely indicate a high Monte Carlo error
with the lower settings. The second issue, the resemblance of the \gls{ABC}
target distribution to the true posterior, is somewhat more difficult to
investigate. We do not have access to the true posterior, even for simulations
where the true parameter values are known. To address this source of error, we
performed marginal parameter estimation with \gls{ABC} by informing
\software{netabc} of some of the true parameter values. Any inaccuracy or bias
observed only in the joint estimation results, but not the marginal estimates, 
is most easily explained by interdependence between parameters in the true
posterior. However, errors observed in both marginal and joint estimates could
be due to either the shape of the true posterior or an inaccurate \gls{ABC}
approximation, and we have no way to distinguish one from the other. In other
words, this experiment provided only an upper bound on the error due to an
inaccurate \gls{ABC} approximation.}

@@

\add{In addition to testing for identifiability, a secondary objective of
this analysis was to validate the use of the tree kernel as a distance measure
for \gls{ABC} in our context. As discussed in \cref{sec:abc}, the choice of
distance function is extremely important for the accuracy of the \gls{ABC}
approximation to the posterior. Therefore, we evaluated two additional tree
statistics in the same manner as we evaluated the tree kernel (that is, by
constructing and testing a classifier). First, we considered Sackin's
index~\autocite{shao1990tree}, which measures the degree of imbalance or
asymmetry in a phylogeny (see \cref{subsec:treeshape}). Sackin's index is
widely used for characterizing phylogenies~\autocite{frost2013modelling} and
has been demonstrated to vary between transmission trees simulated under
different contact network types~\autocite{leventhal2012inferring}. Sackin's
index does not take branch lengths into account, considering only the tree's
topology. The other statistic we considered was the
\gls{nltt}~\autocite{janzen2015approximate}, which compares two trees based on
normalized distributions of their branching times (see
\cref{subsec:treeshape}). In contrast with Sackin's index, the \gls{nltt} does
not explicity consider the trees' topologies, but it does use their normalized
branch lengths. While the \gls{nltt} is a newly developed statistic not yet in
widespread use, the unnormalized \gls{ltt}~\autocite{nee1992tempo} was the
basis of seminal early work extracting epidemiological information from
phylogenies~\autocite{holmes1995revealing}. We expected the tree kernel to
classify the \gls{BA} parameters more accurately than either Sackin's index or
the \gls{nltt}, since the tree kernel takes both topology and branch lengths
into account.}

@@

As in the previous section, we will begin by describing a single experiment,
and then list the variables for which similar experiments were performed. We
varied \gls{alpha} along a narrowly spaced grid of values: 0, 0.01, \ldots, 2.
For each value, fifteen networks were generated with \software{igraph}, and
transmission trees were simulated over each using \software{nettree}. These
trees will be referred to as ``grid trees''. Next, one further test tree was
simulated with the test value \gls{alpha} = 0. Both the grid trees and the test
tree had 500 tips, and were simulated with the other \gls{BA} parameters set to
\add{the known values} $\gls{N} = 5000$, $\gls{m} = 2$, and $\gls{I} = 1000$.
The test tree was compared to each of the grid trees using the tree kernel,
with the meta-parameters set to $\gls{lambda} = 0.3$ and $\gls{sigma} = 4$,
using the \software{treekernel} binary. The median kernel score was calculated
for each grid value, and the scores were normalized such that the area under
the curve was equal to 1. \del{The grid value with the highest median kernel
score was taken as the point estimate for the test value.} \add{For all
parameters except \gls{m}}, 50\% and 95\% highest denstity intervals wer
obtained using the \software{hpd} function in the \software{TeachingDemos}
package~\autocite{snow2013teachingdemos}. \add{Since the \software{hpd}
function assumes a continuous distribution, we implemented our own version for
discrete distributions to use for \gls{m}.}

@@

\add{Our final synthetic data experiment was designed to test the full
\gls{ABC}-\gls{SMC} algorithm by jointly estimating the four parameters of the
\gls{BA} model. We used the standard validation approach of simulating
transmission trees under the model with known parameter values and attempting
to recover those values with \software{netabc}. The algorithm was not informed
of any of the true parameter values for the main set of simulations. Despite
the fact that the parameter values used to generate the simulated transmission
trees were known, the true posterior distributions on the \gls{BA} parameters
were unknown. Therefore, any apparent errors or biases in the estimates could
be due to either poor performance of our method, or to real features of the
posterior distribution. The latter type of error reflects on the suitability of
the model, but does not invalidate the use of our method in cases where the
parameters are more identifiable. Two retrospective experiments were peformed
to disambiguate some of the observed errors: one where we ran a simulation with
increased computational power to test for an increase in accuracy, and a second
where we estimated parameters marginally to remove confounding from the other
parameters in the joint posterior.}

@@

\add{To evaluate the effects of the true parameter values on the accuracy of
the posterior mean estimates, we analysed the \gls{alpha} and \gls{I}
parameters individually using \glspl{GLM}. The response variable was the error
of the point estimate, and the predictor variables were the true values of
\gls{alpha}, \gls{I}, and \gls{m}. We did not test for differences across true
values of \gls{N}, because \gls{N} was not varied in these simulations. The
distribution family and link function for the \glspl{GLM} were chosen as
Gaussian and inverse, respectively, by examination of residual plots and
\gls{AIC}. The $p$-values of the estimated \glspl{GLM} coefficients were
corrected using Holm-Bonferroni correction~\autocite{holm1979simple} with $n =
6$ (two \glspl{GLM} with three predictors each). Because there was clearly
little to no identifiability of \gls{N} and \gls{m} with \gls{ABC} (see results
in next section), we did not construct \glspl{GLM} for those parameters.}

@@

The accuracy of each classifier \add{on all \gls{BA} parameters is shown in}
\del{varied based on the parameter being tested} \cref{fig:rsquared}.
Classifiers based on \del{two other tree statistics,} the \gls{nltt} and
Sackin's index generally exhibited worse performance than the tree kernel,
although the magnitude of the disparity varied between the parameters
(\cref{fig:rsquared}, centre and right). The results were largely robust to
variations in the tree kernel meta-parameters \gls{lambda} and \gls{sigma},
although accuracy varied between different epidemic and sampling scenarios
(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}). \add{For all
parameters except \gls{m}, the absolute number of tips in the tree had a much
greater impact on accuracy than the proportion of infected individuals these
tips represented. However, \gls{m}, both the number and proportion of sampled
tips had a strong impact or the accuracy of the \gls{kSVR}
(\cref{fig:mcrossv}).}

@@

\add{Across all replicates, $R^2$ values for the correlations between the
estimated and true values were
    $\Sexpr{tests["alpha", round(rsquared, 2)]}$,
    $\Sexpr{tests["I", round(rsquared, 2)]}$,
    $\Sexpr{tests["m", round(rsquared, 2)]}$, and
    $\Sexpr{tests["N", round(rsquared, 2)]}$
for \gls{alpha}, \gls{I}, \gls{m}, and \gls{N} respectively. The mean absolute
errors of the point estimates were
    $\Sexpr{tests["alpha", round(abs.err, 2)]}$,
    $\Sexpr{tests["I", round(abs.err)]}$,
    $\Sexpr{tests["m", round(abs.err, 2)]}$, and
    $\Sexpr{tests["N", round(abs.err)]}$,
representing
    $\Sexpr{tests["alpha", round(grid.err)]}$\%,
    $\Sexpr{tests["I", round(grid.err)]}$\%,
    $\Sexpr{tests["m", round(grid.err)]}$\%, and
    $\Sexpr{tests["N", round(grid.err)]}$\%
of the respective grids. For \gls{I} and \gls{N}, relative errors may be more
appropriate to consider. An overestimate of 100 individuals would be very
misleading if the true population was only of size 100, but almost negligable
in a population of size 5000. The mean relative errors for \gls{I} and \gls{N}
respectively were
    $\Sexpr{tests["I", round(rel.err)]}$\% and 
    $\Sexpr{tests["N", round(rel.err)]}$\%.}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{abc-point-estimate-m2}
    \vspace{6pt}
    \caption[
        Posterior mean point estimates for \gls{BA} model parameters obtained
        by running \software{netabc} on simulated data, for simulations with
        $\gls{m} = 2$.
    ]{
        Posterior mean point estimates for \gls{BA} model parameters obtained
        by running \software{netabc} on simulated data, for simulations with
        $\gls{m} = 2$. Dashed lines indicate true values. (A) Estimates of
        \gls{alpha} and \gls{I} which were varied in these simulations against
        known values. (B) Estimates of \gls{m} and \gls{N} which were held
        fixed in these simulations at the values $\gls{m} = 2$ and 
        $\gls{N} = 5000$.
    }
    \label{fig:abcptm2}
\end{figure}

@@

\add{To analyse the effects of the true parameter values on the
accuracy our estimates of \gls{alpha} and \gls{I}, we fitted one \glspl{GLM}
for each of these two parameters, with error rate as the dependent variable and
the true parameter values as independent variables. Since the estimates of
\gls{m} and \gls{N} were roughly equal across all simulations
(\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}), \glspl{GLM} were not fitted for
these parameters. The estimated coefficients are shown in
\cref{tab:glmalpha,tab:glmI}. For the parameters \gls{alpha}, \gls{I}, and
\gls{N}, the \gls{GLM} were fitted using the inverse link function. That is, if
$p$ is the true value of the parameter and $\hat{p}$ is a random variable
representing our estimate of the parameter, the \gls{GLM} posits a relationship
of the form}
\[
    \madd{\E(|p-\hat{p}|) = (\beta_0 + \beta_\alpha \alpha + \beta_I I + \beta_m m)^{-1}},
\]
\add{where the $\beta$'s are coefficients to be fitted. If the true value
\gls{alpha}, say, is increased by one, the \emph{inverse} of the expected
absolute error will increase by $\beta_\alpha$. If $\beta_\alpha$ is positive,
it means that the absolute error decreases as the true value of \gls{alpha}
increases.}

>>>>

As described in \cref{sec:smc}, \software{netabc} keeps track of a population
of particles $x^{(k)}$ indexed by an integer $k$, each of which contains
particular parameter values $\theta^{(k)}$ for the \add{contact network} model
we are trying to fit \add{to the input tree}. A small number \gls{M} of contact
networks $z^{(k,i)}$, $1 \leq i \leq \gls{M}$, are generated under the model
for each particle in accordance with that particle's parameters. An epidemic is
simulated over each of these networks using Gillespie simulation, and by
keeping track of its progress, a transmission tree is obtained. Thus, each
particle becomes associated with several simulated transmission trees. These
trees are compared to the input tree using the tree kernel. Particles are
weighted according to the similarity of their associated simulated trees with
the true tree, with more similar trees receiving higher weights. The particles
are iteratively perturbed to explore the parameter space, and particles with
simulated trees too distant from the true tree are periodically dropped and
resampled. Once a convergence criterion is attained, the final set of particles
is used as a Monte Carlo approximation to the target distribution of \gls{ABC},
which is assumed to resemble the posterior distribution on model parameters
(see \cref{sec:abc}). A graphical schematic of this algorithm is shown in
\cref{fig:abcsmc}.

@@

\add{We present here a non-rigorous justification for the use of
\gls{ABC} for the problem at hand, as opposed to more frequently-used
approaches for fitting mathematical models (see \cref{chp:prelim}). Consider a
contact network model with parameters $\theta$, and an estimated transmission
tree \gls{T}. Taking a Bayesian approach, our aim is to obtain a sample from
the posterior distribution on the model's parameters given our data,}
\[
    \madd{\post(\theta \mid \gls{T}) = \frac{\lik(\gls{T} \mid \theta) \prior(\theta)}
    {\int_{\Theta}\lik(\gls{T} \mid \theta) \prior(\theta) \d\theta}.}
\]
\add{For all but the simplest models, the normalizing constant in the
denominator is an intractable integral. What we shall argue here is that, in
contrast to most commonly studied mathematical models, the likelihood
$\lik(\gls{T} \mid \theta)$ is also likely to be intractable in our case.}

@@

\add{We have shown that both the normalizing constant $\int_\Theta \lik(\gls{T}
\mid \theta) \prior(\theta) \d\theta$ and the likelihood $\lik(T \mid \theta)$
are likely computationally prohibitive to calculate. If this is the case, the
problem of fitting contact network models to phylogenies seems to be of the
\emph{doubly-intractable} type~\autocite{murray2012mcmc}, which would imply that
these models are not amenable to neither \gls{ML} nor Bayesian inference
techniques. Although both methods are able to cope with an intractable
normalizing constant (for example, by local search for \gls{ML} or Bayesian
\gls{MCMC}), neither can avoid the intractable likelihood calculations. This
justifies the use of \gls{ABC}, which is a likelihood-free method.}

@@

\add{From a public health standpoint, all four parameters are of some interest.
The prevalence \gls{I} can be used to estimate the resources required to combat
an ongoing epidemic, while total susceptible population size \gls{N} provides a
similar metric for preventative measures. The average degree of the network, in
this case $2\gls{m}$, is directly related to \gls{R0}, the basic reproductive
number~\autocite{britton2002bayesian}. \gls{R0} quantifies the average number
of secondary infections ultimately caused by one infected individual; higher
\gls{R0} generally indicates faster epidemic growth and/or larger eventual
epidemic size~\autocite{anderson1992infectious}. In a homogeneously mixed
population, the theoretically expected proportion of the population which must
be vaccinated to control an epidemic (the \defn{vaccination threshold}) can be
expressed in terms of \gls{R0}~\autocite{anderson1982directly}. Although
optimal vaccination strategies may differ in heterogeneous contact
structures~\autocite{ma2013importance}, it is reasonable to suppose that there
would still be a relationship between \gls{m}, \gls{R0}, and the vaccination
threshold. The preferential attachment power \gls{alpha} quantifies the degree
to which high-degree nodes, also called
superspreaders~\autocite{kemper1980identification}, characterize the network
structure. Superspreaders have been hypothesized to play a disproportionately
greater role in the spread of several diseases~\autocite{stadler2013uncovering,
shen2004superspreading}. If so, network-based
interventions~\autocite{little2014using, wang2015targeting} may be worth
considering as part of an epidemic control strategy. \gls{alpha} can also offer
some insight into how the network would react to the removal of nodes.
\textcite{dombrowski2013topological} found evidence of preferential attachment
in \gls{IDU} networks, and suggested that as a consequence of this
characteristic, the removal of random nodes (such as through a police
crackdown) might inadvertently make it easier for epidemics to spread. When
individuals with only one or two connections lose them, they might tend to seek
out well-known (that is, high-degree) members of the community, thus increasing
those individuals' connectivity even further. Although we do not consider a 
dynamic network in this work, it is not unreasonable that the network could be
close to static over short periods of time, and this static structure might be
informative of future dynamic behaviour.}

\add{Though it is theoretically possible to estimate all four \gls{BA}
parameters using more conventional approaches, the cost of doing so may be
high, in addition to other situation-specific challenges. In theory, any
network parameter can be estimated by explicitly constructing the contact
network, although this is highly resource intensive and is hampered by
misreporting and other challenges~\autocite{eames2015six}. All parameters
become more difficult to estimate when the infected population is ``hidden''
due to illegal or stigmatized behaviour, as is sometimes the case with
\gls{HIV} outbreaks among \gls{IDU}, \gls{MSM}, or sex workers. In such cases,
phylodynamic methods may offer the advantage of providing information about the
whole population from only a small sample. A survey, for example, will not tell
us if there are large groups of the population that we simply have not sampled.
Our hope is that estimating \gls{N} and \gls{I} phylogenetically might provide
this additional information. The average degree of the network, $2\gls{m}$, is
also estimable by a survey, although individuals may be unwilling or unable to
disclose how many contacts they have had. Sequence data also provide an
advantage in this respect -- they are objective and do not share the same
biases as self-reported partner counts. The estimation of \gls{alpha} is more
complex, as this parameter is most strongly reflected in the connectivity of
very high degree nodes, who are rare in the population. Locating them might
require contact tracing, or respondent-driven
sampling~\autocite{heckathorn1997respondent}. Even if the full degree
distribution of a network is available, there are models other than
preferential attachment which can produce scale-free
networks~\autocite[\eg][]{kumar2000stochastic}. \textcite{de2007preferential}
were able to estimate \gls{alpha} by maximum likelihood using partner count
data from several sequential time intervals, but they admit such detailed data
are not usually available. Moreover, their dataset was constructed via a random
survey, which would likely miss the few high-degree nodes characterizing a
power law degree distribution. In summary, each of the \gls{BA} parameters may
be estimated without phylodynamics, but there are sufficient difficulties that
we believe an alternative method using sequence data is warranted.}

@@

\add{A second experiment was designed to test whether we could actually
estimate the parameters numerically, rather than just telling three different
values apart, and also to assess how identifiability varied in different
regions of the parameter space. An individual tree was compared to simulated
trees on a one-dimensional grid of values of one \gls{BA} parameter, to obtain
a ``distribution'' of kernel score values. From this distribution, estimates
and credible intervals of the parameter could be calculated. Repeating this
experiment with trees located throughout the parameter space allowed us to
better quantify the identifiability. Furthermore, doing marginal estimation
(that is, estimating one parameter with all others fixed) can provide insight
into any biases observed when doing joint estimation with \gls{ABC}. If grid
search is inaccurate, it indicates a lack of parameter identifiability.
However, if the marginal grid search estimates are accurate but the estimates
obtained with \gls{ABC} are biased, this points to confounding between the
parameters which could only be observed when they are all estimated jointly.}

@@

\add{The Monte Carlo approximation error is fairly easily quantified by simply
increasing the computing power used for \gls{SMC}. We ran one simulation using
a larger number of particles, more simulated datasets per particle, and a
higher value for \gls{alphaess}. A substantial improvement in accuracy
resulting from these changes would likely indicate a high Monte Carlo error
with the lower settings. The second issue, the resemblance of the \gls{ABC}
target distribution to the true posterior, is somewhat more difficult to
investigate. We do not have access to the true posterior, even for simulations
where the true parameter values are known. To address this source of error, we
performed marginal parameter estimation with \gls{ABC} by informing
\software{netabc} of some of the true parameter values. Any inaccuracy or bias
observed only in the joint estimation results, but not the marginal estimates, 
is most easily explained by interdependence between parameters in the true
posterior. However, errors observed in both marginal and joint estimates could
be due to either the shape of the true posterior or an inaccurate \gls{ABC}
approximation, and we have no way to distinguish one from the other. In other
words, this experiment provided only an upper bound on the error due to an
inaccurate \gls{ABC} approximation.}

@@

\add{In addition to testing for identifiability, a secondary objective of
this analysis was to validate the use of the tree kernel as a distance measure
for \gls{ABC} in our context. As discussed in \cref{sec:abc}, the choice of
distance function is extremely important for the accuracy of the \gls{ABC}
approximation to the posterior. Therefore, we evaluated two additional tree
statistics in the same manner as we evaluated the tree kernel (that is, by
constructing and testing a classifier). First, we considered Sackin's
index~\autocite{shao1990tree}, which measures the degree of imbalance or
asymmetry in a phylogeny (see \cref{subsec:treeshape}). Sackin's index is
widely used for characterizing phylogenies~\autocite{frost2013modelling} and
has been demonstrated to vary between transmission trees simulated under
different contact network types~\autocite{leventhal2012inferring}. Sackin's
index does not take branch lengths into account, considering only the tree's
topology. The other statistic we considered was the
\gls{nltt}~\autocite{janzen2015approximate}, which compares two trees based on
normalized distributions of their branching times (see
\cref{subsec:treeshape}). In contrast with Sackin's index, the \gls{nltt} does
not explicitly consider the trees' topologies, but it does use their normalized
branch lengths. While the \gls{nltt} is a newly developed statistic not yet in
widespread use, the unnormalized \gls{ltt}~\autocite{nee1992tempo} was the
basis of seminal early work extracting epidemiological information from
phylogenies~\autocite{holmes1995revealing}. We expected the tree kernel to
classify the \gls{BA} parameters more accurately than either Sackin's index or
the \gls{nltt}, since the tree kernel takes both topology and branch lengths
into account.}

@@

As in the previous section, we will begin by describing a single experiment,
and then list the variables for which similar experiments were performed. We
varied \gls{alpha} along a narrowly spaced grid of values: 0, 0.01, \ldots, 2.
For each value, fifteen networks were generated with \software{igraph}, and
transmission trees were simulated over each using \software{nettree}. These
trees will be referred to as ``grid trees''. Next, one further test tree was
simulated with the test value \gls{alpha} = 0. Both the grid trees and the test
tree had 500 tips, and were simulated with the other \gls{BA} parameters set to
\add{the known values} $\gls{N} = 5000$, $\gls{m} = 2$, and $\gls{I} = 1000$.
The test tree was compared to each of the grid trees using the tree kernel,
with the meta-parameters set to $\gls{lambda} = 0.3$ and $\gls{sigma} = 4$,
using the \software{treekernel} binary. The median kernel score was calculated
for each grid value, and the scores were normalized such that the area under
the curve was equal to 1. \del{The grid value with the highest median kernel
score was taken as the point estimate for the test value.} \add{For all
parameters except \gls{m}}, 50\% and 95\% highest density intervals were
obtained using the \software{hpd} function in the \software{TeachingDemos}
package~\autocite{snow2013teachingdemos}. \add{Since the \software{hpd}
function assumes a continuous distribution, we implemented our own version for
discrete distributions to use for \gls{m}.}

@@

\add{Our final synthetic data experiment was designed to test the full
\gls{ABC}-\gls{SMC} algorithm by jointly estimating the four parameters of the
\gls{BA} model. We used the standard validation approach of simulating
transmission trees under the model with known parameter values and attempting
to recover those values with \software{netabc}. The algorithm was not informed
of any of the true parameter values for the main set of simulations. Despite
the fact that the parameter values used to generate the simulated transmission
trees were known, the true posterior distributions on the \gls{BA} parameters
were unknown. Therefore, any apparent errors or biases in the estimates could
be due to either poor performance of our method, or to real features of the
posterior distribution. The latter type of error reflects on the suitability of
the model, but does not invalidate the use of our method in cases where the
parameters are more identifiable. Two retrospective experiments were performed
to disambiguate some of the observed errors: one where we ran a simulation with
increased computational power to test for an increase in accuracy, and a second
where we estimated parameters marginally to remove confounding from the other
parameters in the joint posterior.}

@@

\add{To evaluate the effects of the true parameter values on the accuracy of
the posterior mean estimates, we analyzed the \gls{alpha} and \gls{I}
parameters individually using \glspl{GLM}. The response variable was the error
of the point estimate, and the predictor variables were the true values of
\gls{alpha}, \gls{I}, and \gls{m}. We did not test for differences across true
values of \gls{N}, because \gls{N} was not varied in these simulations. The
distribution family and link function for the \glspl{GLM} were chosen as
Gaussian and inverse, respectively, by examination of residual plots and
\gls{AIC}. The $p$-values of the estimated \glspl{GLM} coefficients were
corrected using Holm-Bonferroni correction~\autocite{holm1979simple} with $n =
6$ (two \glspl{GLM} with three predictors each). Because there was clearly
little to no identifiability of \gls{N} and \gls{m} with \gls{ABC} (see results
in next section), we did not construct \glspl{GLM} for those parameters.}

@@

The accuracy of each classifier \add{on all \gls{BA} parameters is shown in}
\del{varied based on the parameter being tested} \cref{fig:rsquared}.
Classifiers based on \del{two other tree statistics,} the \gls{nltt} and
Sackin's index generally exhibited worse performance than the tree kernel,
although the magnitude of the disparity varied between the parameters
(\cref{fig:rsquared}, centre and right). The results were largely robust to
variations in the tree kernel meta-parameters \gls{lambda} and \gls{sigma},
although accuracy varied between different epidemic and sampling scenarios
(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}). \add{For all
parameters except \gls{m}, the absolute number of tips in the tree had a much
greater impact on accuracy than the proportion of infected individuals these
tips represented. However, for \gls{m}, both the number and proportion of
sampled tips had a strong impact or the accuracy of the \gls{kSVR}
(\cref{fig:mcrossv}).}

@@

\add{Across all replicates, $R^2$ values for the correlations between the
estimated and true values were
    $\Sexpr{tests["alpha", round(rsquared, 2)]}$,
    $\Sexpr{tests["I", round(rsquared, 2)]}$,
    $\Sexpr{tests["m", round(rsquared, 2)]}$, and
    $\Sexpr{tests["N", round(rsquared, 2)]}$
for \gls{alpha}, \gls{I}, \gls{m}, and \gls{N} respectively. The mean absolute
errors of the point estimates were
    $\Sexpr{tests["alpha", round(abs.err, 2)]}$,
    $\Sexpr{tests["I", round(abs.err)]}$,
    $\Sexpr{tests["m", round(abs.err, 2)]}$, and
    $\Sexpr{tests["N", round(abs.err)]}$,
representing
    $\Sexpr{tests["alpha", round(grid.err)]}$\%,
    $\Sexpr{tests["I", round(grid.err)]}$\%,
    $\Sexpr{tests["m", round(grid.err)]}$\%, and
    $\Sexpr{tests["N", round(grid.err)]}$\%
of the respective grids. For \gls{I} and \gls{N}, relative errors may be more
appropriate to consider. An overestimate of 100 individuals would be very
misleading if the true population was only of size 100, but almost negligible
in a population of size 5000. The mean relative errors for \gls{I} and \gls{N}
respectively were
    $\Sexpr{tests["I", round(rel.err)]}$\% and 
    $\Sexpr{tests["N", round(rel.err)]}$\%.}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{abc-point-estimate-m2}
    \caption[
        Posterior mean point estimates for \gls{BA} model parameters obtained
        by running \software{netabc} on simulated data, for simulations with
        $\gls{m} = 2$.
    ]{
        Posterior mean point estimates for \gls{BA} model parameters obtained
        by running \software{netabc} on simulated data, for simulations with
        $\gls{m} = 2$. Dashed lines indicate true values. (A) Estimates of
        \gls{alpha} and \gls{I} which were varied in these simulations against
        known values. (B) Estimates of \gls{m} and \gls{N} which were held
        fixed in these simulations at the values $\gls{m} = 2$ and 
        $\gls{N} = 5000$.
    }
    \label{fig:abcptm2}
\end{figure}

@@

\add{To analyze the effects of the true parameter values on the
accuracy our estimates of \gls{alpha} and \gls{I}, we fitted one \glspl{GLM}
for each of these two parameters, with error rate as the dependent variable and
the true parameter values as independent variables. Since the estimates of
\gls{m} and \gls{N} were roughly equal across all simulations
(\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}), \glspl{GLM} were not fitted for
these parameters. The estimated coefficients are shown in
\cref{tab:glmalpha,tab:glmI}. For the parameters \gls{alpha}, \gls{I}, and
\gls{N}, the \gls{GLM} were fitted using the inverse link function. That is, if
$p$ is the true value of the parameter and $\hat{p}$ is a random variable
representing our estimate of the parameter, the \gls{GLM} posits a relationship
of the form}
\[
    \madd{\E(|p-\hat{p}|) = (\beta_0 + \beta_\alpha \alpha + \beta_I I + \beta_m m)^{-1}},
\]
\add{where the $\beta$'s are coefficients to be fitted. If the true value
\gls{alpha}, say, is increased by one, the \emph{inverse} of the expected
absolute error will increase by $\beta_\alpha$. If $\beta_\alpha$ is positive,
it means that the absolute error decreases as the true value of \gls{alpha}
increases.}
